Rectangulotopes

### 2024-04-26 draft/rectangulotopes.tex
add ref

The combinatorics of rectangulations has been studied for nearly two decades~\cite{MR2233287,MR2763051,MR2871762,MR2864445,MR3084577,MR3192492,MR3878132,MR4598046,AB24,ACFF24}.
In fact, several combinatorial ideas can even be traced back to the problem of ``squaring the square'' studied since the 1930s~\cite{A30,Abe32,MR0000470,MR0003040}. Also, some results were initially motivated by applications to the design of \defn{floorplans} for very large scale integrated circuits, and published in electrical engineering journals, see for instance Z. C. Shen and C. C. N. Chu~\cite{SC03}, and R. Fujimaki, Y. Inoue, and T. Takahashi~\cite{FT07,TF08,ITF09,FIT09}.

@@

\begin{lemma}
  \label{lem:lodaymax}
  Let $\sigma \in \f{S}_n$ and $I$ be an interval of~$[n]$ of size at least~$2$.
  The unique vertex of the up shard polytope~$\SP(\upArc{I})$ (resp.~down shard polytope~$\SP(\downArc{I})$) that is extreme with respect to the direction $\sigma^{-1}$ is $\b{e}_i$ (resp.~$-\b{e}_i$), where $i$ is the maximum (resp. the minimum) of~$\sigma^{-1}$ in $I$.
\end{lemma}

Note that this extremal vertex should only depend on the weak rectangulation congruence class of $\sigma$.
This, in passing, gives us an alternative characterization of the weak rectangulation congruence:
$\sigma \weakeq \tau$ if and only if for any interval $I$ of $[n]$,
$\arg\max_{j\in I} \sigma^{-1}_j=\arg\max_{j\in I} \tau^{-1}_j$, and
$\arg\min_{j\in I} \sigma^{-1}_j=\arg\min_{j\in I} \tau^{-1}_j$.

@@

\begin{lemma}
  \label{lem:yinminmax}
  Let $\sigma\in\f{S}_n$, let~$I$ and~$J$ be two consecutive intervals of~$[n]$, and let $i \eqdef \arg\max_{k\in I} \sigma^{-1}_k$ and $j \eqdef \arg\min_{k\in J} \sigma^{-1}_k$.
  The unique vertex of the yin shard polytope~$\SP(\yinArc{I}{J})$ that is extreme with respect to the direction $\sigma^{-1}$ is $\b{e}_i-\b{e}_j$ if $\sigma^{-1}_i > \sigma^{-1}_j$, and $\b{0}$ otherwise.
\end{lemma}

Note again that by definition, this extremal vertex should only depend on the strong congruence class of $\sigma$.

>>>>

The combinatorics of rectangulations has been studied for nearly two decades~\cite{MR2233287,MR2763051,MR2871762,MR2864445,MR3084577,MR3192492,MR3878132,MR4598046,MR4667583,AB24,ACFF24}.
In fact, several combinatorial ideas can even be traced back to the problem of ``squaring the square'' studied since the 1930s~\cite{A30,Abe32,MR0000470,MR0003040}. Also, some results were initially motivated by applications to the design of \defn{floorplans} for very large scale integrated circuits, and published in electrical engineering journals, see for instance Z. C. Shen and C. C. N. Chu~\cite{SC03}, and R. Fujimaki, Y. Inoue, and T. Takahashi~\cite{FT07,TF08,ITF09,FIT09}.

@@

\begin{lemma}
  \label{lem:lodaymax}
  Let $\sigma \in \f{S}_n$ and $I$ be an interval of~$[n]$ of size at least~$2$.
  The unique vertex of the up shard polytope~$\SP(\upArc{I})$ (resp.~down shard polytope~$\SP(\downArc{I})$) that is extreme with respect to the direction $\sigma^{-1}$ is $\b{e}_i$ (resp.~$-\b{e}_i$), where $i$ is the maximum (resp.~the minimum) of~$\sigma^{-1}$ in $I$.
\end{lemma}

Note that this extremal vertex should only depend on the weak rectangulation congruence class of $\sigma$.
This, in passing, gives us alternative characterizations of the sylvester congruence~$\frown$, the anti-sylvester congruence~$\smile$, and the weak rectangulation congruence $\weakeq$.
We have~$\sigma \frown \tau$ if and only if $\arg\max_{j\in I} \sigma^{-1}_j=\arg\max_{j\in I} \tau^{-1}_j$ for any interval $I$ of $[n]$.
Similarly, $\sigma \smile \tau$ if and only if $\arg\min_{j\in I} \sigma^{-1}_j=\arg\min_{j\in I} \tau^{-1}_j$ for any interval $I$ of $[n]$.
Finally, $\sigma\weakeq\tau$ if both conditions hold.

@@

\begin{lemma}
  \label{lem:yinminmax}
  Let $\sigma\in\f{S}_n$, let~$I$ and~$J$ be two consecutive intervals of~$[n]$, and let $i\eqdef \arg\max_{k\in I} \sigma^{-1}_k$ and $j \eqdef \arg\min_{k\in J} \sigma^{-1}_k$ (resp.~$i\eqdef \arg\min_{k\in I} \sigma^{-1}_k$ and $j \eqdef \arg\max_{k\in J} \sigma^{-1}_k$).
  The unique vertex of the yin (resp.~yang) shard polytope~$\SP(\yinArc{I}{J})$ (resp.~$\SP(\yangArc{I}{J})$) that is extreme with respect to the direction $\sigma^{-1}$ is $\b{e}_i-\b{e}_j$ (resp.~$\b{e}_j-\b{e}_i$) if $\sigma^{-1}_i > \sigma^{-1}_j$ (resp.~$\sigma^{-1}_i < \sigma^{-1}_j$), and $\b{0}$ otherwise.
\end{lemma}

Note again that by definition, this extremal vertex should only depend on the strong congruence class of $\sigma$.
This, again, gives us alternative characterizations of the yin congruence $\sim$, the yang congruence $\backsim$, and the strong rectangulation congruence $\strongeq$.
We have $\sigma \sim \tau$ if and only if for any pair of consecutive intervals $I$ and $J$ of $[n]$, 
$\arg\max_{k\in I} \sigma^{-1}_k = \arg\max_{k\in I} \tau^{-1}_k$ and 
$\arg\min_{k\in J} \sigma^{-1}_k = \arg\min_{k\in J} \tau^{-1}_k$ (that is, if $\sigma \weakeq \tau$),
and, if we let $i$ and $j$ be defined as in Lemma~\ref{lem:yinminmax} as the indices of the maximum in $I$ and the minimum in $J$,
$\sigma^{-1}_i > \sigma^{-1}_j \Leftrightarrow \tau^{-1}_i > \tau^{-1}_j$.
A similar statement holds for the yang congruence $\backsim$, by exchanging $\min$ and $\max$.
Finally, $\sigma\strongeq\tau$ if both sets of conditions hold.
